Towards a classification framework for social machines

### 2013-02-26 latex/paper_new.tex
written sec
intermediary version

\section{Social machines: an introduction}
Once upon a time 'machines' were programmed by programmers and used by users. The success of the Web has changed this relationship: we now see configurations of people interacting with content and with each other, typified by social Web sites. Rather than drawing a line through such Web-based systems to separate the human and machine parts (as Computer Science has traditionally done), we can now draw a line around them  and  treat  each  such  assembly  as  a  'social  machine'  -  a  machine  in which  digital  and  human components are seamlessly interwoven. This crucial transition in thinking acknowledges the reality of current socio-technical systems, and is essential to underpin any understanding of the science and engineering of their future development towards pervasive ecosystems of co-evolving social machines.

A number of prominent examples aside, today we still see a divide between conventional IT systems dedicated to data- and computation-intensive tasks, and Web $2.0$ sites offering some combination of well-known participatory features, in which user-generated content and the underlying social network evolve dynamically and hand-in-hand. As technology becomes more and more ubiquitous, many of the challenges we will witness in any domain of our life, business, and society will need solutions that rely on both of these axes: a sophisticated  combination  of  data-intensive, complex automation and deep community involvement. This suggests the need for new types of systems to tackle these challenges, and to achieve requires a through understanding of the science and engineering of social machines.

As a first step, we propose a classification framework for social machines, which can be used to describe, study, and compare this expanding field of interdisciplinary research and the increasing number of systems combining human and computational intelligence to serve purposes as diverse as traffic management, healthcare, media and journalism, or design and natural sciences.

In this paper we introduce the current version of the framework, which is the result of a knowledge elicitation process performed internally in our lab. Our aim is to present it to the broader community and extend and revise it based on their feedback in order to achieve a shared understanding, terminology, and set of descriptive features to enable a systematic study of existing and future social machines. Using a common descriptive classification allows researchers to observe the effects of specific technical properties of a system on social behavior, identify design and evolution patterns, learn about social network formation and dynamics, and devise incentive mechanisms to encourage large-scale participation. A second audience group targeted by our work are developers of social machines. For them, the framework may provide a theoretical grounding for the realization of new systems, informing system design and operation in terms of critical features and predicted user behavior and community development.

\section{What are social machines?}
Social machines can be basically characterized as assemblies of human and automatic components and their interaction. A  traditional  database,  with  users  looking  up  records  independently  of  each  other  and  an  administrator responsible for content, has some of the right ingredients but there is really no social component in this strict provider-consumer relationship. When we fill out a form (e.g., health information, birds spotted in my garden) it  is minimally  a  social machine  because humans  are  part  of  the  'social  computation'.  This  becomes  richer when  the database  is  curated by members  of  the  community  (e.g., Wikipedia) and when  the  social network adds value implicitly (Amazon) or explicitly (Facebook). We anticipate many categories of social machine in this continuum, and they may manifest themselves as Web site or app, or indeed in the Internet of Things.

\section{Comparison with related fields}

A Social Machine includes human and machine two major components, therefore the study of social machine is an interdisciplinary endeavour. Since the Turning machine was invented, there has been a `paradigm shift' \cite{kuhn1996structure} in using computers, where the computer facilitates human communication rather acts as a purely calculative device. The Computer-supported Collaboration Work (CSCW) \cite{grudin1994computer} emerged in 1984 represents such shift. It addresses ``how collaborative activities and their coordination can be supported by means of computer systems''. This concept later evolved into a broader sense of  Computer-supported collaboration (CSC). The Social computing \cite{parameswaran2007research} is another shift, which refers systems that support ``the gathering, representation, processing, use, and dissemination of information that is distributed across social collectivities such as teams, communities, organizations, and markets''. To compare with the CSC, the social computing has more focus in `computing' rather than `collaboration'. Similarly, the Human Computation \cite{von2009human} is another paradigm towards the computational direction, which utilize human processing power to solve problems that computers cannot yet solve. All these three mentioned fields overlap in their concepts and are mainly popularised in the computer science. In the Web 2.0 era, there were few generic concepts/buzzword emerged on the idea of utilising the human power which are not only used in computer science, but also in other fields such as social science, business etc. Crowding sourcing \cite{howe2006crowdsourcing} as an example, describes the idea that taking a job traditionally performed by a designated agent and outsourcing it to an undefined, generally large group of people in the form of an open call. Open Innovation \cite{chesbrough2008open} advocates the similar idea with empathizing on the innovation in firms. Surowiecki also argued in the book of the wisdom of crowds \cite{surowiecki2005wisdom} that the aggregation of information in groups, resulting in decisions that are often better than could have been made by any single member of group. Collective Intelligence described by Malone \cite{malone2009harnessing}, defines an even more broadly concept that ``groups of individuals doing things collectively that seem intelligent.''
The concept overlapping between human computation, crowdsourcing, social computing and collective intelligence is discussed in \cite{quinn2011human}.
In  social machines  the  social and  the  technical are equal  partners. While  crowdsourcing  could  be  viewed  as  'human  subroutines'  and  'citizen  sensors',  the social machine perspective is one of incentivised participation and not humans enslaved to computers. We define the scope of social machine in a venn digram in Figure \ref{socialmachine}.

The study of social machines is naturally an interdisciplinary endeavor. Since the Turning machine was invented, we have experienced a paradigm shift in computer science from computers as purely calculative devices into a worldwide interconnected computer infrastructure which facilitates almost any type of human endeavor. The Computer-supported Collaboration Work (CSCW) \cite{grudin1994computer} emerged in 1984 represents such shift. It addresses ``how collaborative activities and their coordination can be supported by means of computer systems''. This concept later evolved into a broader sense of  Computer-supported collaboration (CSC). The Social computing \cite{parameswaran2007research} is another shift, which refers systems that support ``the gathering, representation, processing, use, and dissemination of information that is distributed across social collectivities such as teams, communities, organizations, and markets''. To compare with the CSC, the social computing has more focus in `computing' rather than `collaboration'. Similarly, the Human Computation \cite{von2009human} is another paradigm towards the computational direction, which utilize human processing power to solve problems that computers cannot yet solve. All these three mentioned fields overlap in their concepts and are mainly popularised in the computer science. In the Web 2.0 era, there were few generic concepts/buzzword emerged on the idea of utilising the human power which are not only used in computer science, but also in other fields such as social science, business etc. Crowding sourcing \cite{howe2006crowdsourcing} as an example, describes the idea that taking a job traditionally performed by a designated agent and outsourcing it to an undefined, generally large group of people in the form of an open call. Open Innovation \cite{chesbrough2008open} advocates the similar idea with empathizing on the innovation in firms. Surowiecki also argued in the book of the wisdom of crowds \cite{surowiecki2005wisdom} that the aggregation of information in groups, resulting in decisions that are often better than could have been made by any single member of group. Collective Intelligence described by Malone \cite{malone2009harnessing}, defines an even more broadly concept that ``groups of individuals doing things collectively that seem intelligent.''
The concept overlapping between human computation, crowdsourcing, social computing and collective intelligence is discussed in \cite{quinn2011human}.
In  social machines  the  social and  the  technical are equal  partners. While  crowdsourcing  could  be  viewed  as  'human  subroutines'  and  'citizen  sensors',  the social machine perspective is one of incentivised participation and not humans enslaved to computers. We define the scope of social machine in a conceptual digram in Figure \ref{socialmachine}.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=8cm]{img/socialmachinescope.png}
\caption{Social machines and related areas} \label{social machine}
\end{center}
\end{figure}

\section{Classification of social machines}

\subsection{Methodology} \label{meth}

The classification framework presented in this paper was created through knowledge elicitation ~\cite{knowledgeelicitation}. In particular we used the {\it repertory grid} elicitation
technique~\cite{kelly} in order to derive an initial set of {\it elements}, which represent instances of social machines, and {\it constructs}, which capture their most important characteristics.

In repertory grid elicitation a software tool is used to ask users to describe constructs
that differentiate between elements - for example,  prompting the user to create a construct that
differentiates between {\it GalaxyZoo} and {\it Facebook}, as two prominent example of social machines our classification framework should apply to. The user describes the opposing poles of the construct - in this case, the user may decide on 'For Science' and 'For Socialising' in order to capture the core distinction between the two system. The user then rates every element with value from $1$ to $5$ on this construct, where $1$ represents an element that is purely
'For Science', and $5$ one that mainly serves  'socialising' purposes. Elements can also be rated with values between the poles if they may correspond to both or neither. The repertory grid software then identifies elements that are the least differentiated when requesting new constructs, in an attempt to elicit as much knowledge as possible from the user. We applied this technique to create an initial set of examples of social machines and to classify them according to specific features. To do so, we asked then computer science researchers familiar with the field to create their own repertory grids, and populate the elements from their own knowledge, and created the constructs using the standard repertory grid elicitation technique described above. This exercise led to $10$ grids, the union of which comprised a total of $56$ unique elements (social machines)
and $117$ different constructs (classifying factors). As the aim of this initial phase was to understand how people perceive the notion of social machine and their most distinctive properties, we allowed the participants to choose the social machines they are familiar with and describe them in their own terms.

@@

While determining the intersection of elements was straightforward, the consolidation of the
constructs required a more thorough process. We embarked on a manual process of grouping the constructs into rough clusters, based around the areas they cover. For example, we used construct
clusters including: Users, Motivation, Popularity, Technology, and Purpose. We then examined
each construct to determine which were equivalent, and whether we could re-word or subsume
existing constructs to cover the same aspects as a construct, with the aim to remove any
redundancy, to cut the list of constructs down to a manageable number, but to ensure that
all of the constructs that were elicited were somehow represented in the final set. This
process involved four of the authors discussing the choices, and finally agreeing on a
consolidated set of $31$ constructs.

%Even with this reduced set of constructs, if a single person were to classify all 56
%elements, this would still require 1736 individual decisions, and hence is unfeasible to
%expect full coverage of the elements using all of these constructs, despite how representative
%they are.
%
%[something here about doing 2 grids each and verifying the results?]
%
%To extract greater usefulness, and the ability to use our constructs to classify social
%machines we re-evaluated the constructs and the semantics that they represent into richer
%constructs that are not constrained by the 1-5 rating of repertory grid constructs.

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{|p{8cm}|}
\hline
{\bf Users and Roles} \\
\hline
Generality of Audience \\
Participant autonomy \\
Participant anonymity \\
Extent of hierarchical organisation of roles \\
Clear separation of roles/responsibilities among participants \\
\hline
{\bf Tasks} \\
\hline
Activities involve creative production of content \\
Activities involve subjective appraisal of content \\
Activities involve solving (a definable) computation task or set of problems\\
Tasks are domain-specific \\
The machine owner (service/platform) derives benefit from participation \\
Activities and tasks are pre-defined or participant-defined \\
Variation in types of contributions and tasks \\
\hline
{\bf Motivation} \\
\hline
Participants are intrinsically motivated:\\
\hspace{1cm} to gain/share knowledge \\
\hspace{1cm} to ``get something done'' \\
\hspace{1cm} to ``for fun/entertainment''\\
\hspace{1cm} to ``be social'' \\
\hspace{1cm} for the benefit of a specific group of people who need help \\ 
\hspace{1cm} for the benefit of society as a whole \\ 
Participation is motivated by extrinsic reward (payment, status etc) \\
\hline
{\bf Context of Participation} \\
\hline
Participants' participate to fulfill needs of a role  \\
\hspace{1cm} work-related/professional role
\hspace{1cm} home/family related
\hspace{1cm} social role
\hspace{1cm} leisure/entertainment role
Participants' physical location is relevant to the service \\
Participation is done via: \\
\hspace{1cm} mobile devices \\
\hspace{1cm} web browsers \\
\hspace{1cm} apps \\
\hspace{1cm} sensors/sensing (location sensing and wearable devices) \\
Participation is done in a mobile context \\
\hline
\end{tabular}
\end{center}
\caption{Consolidated constructs of social machines.} \label{table:constructs}
\end{table}

@@

\subsubsection{Tasks, Problems and Activiites}
The first set of constructs pertain to \emph{what} the participants
do, both individually within the system, and collectively as a whole.
Individual activities might include creation of content, subjective
appraisal of existing content, posing of problems, solving of
problems, andf so on.  The resulting computation, looked at from a
macroscopic whole, might be the identification of high-quality
artistic content, insight, or creative works, collective problem
solving, and so on.  Such tasks might be pre-defined by the system
designers, or brought to the system by participants.  Futher, these
tasks might be of a specific type or class, or of many varied types or
encompass a general class of activities.  To capture these
difeferences, we include a construct that differentiates machines with
pre-defined problem spaces or are open to participant-provided tasks
and activities.  Finally, we ask whether the benefit derived by
participants is distinct from that gained by machine/platform owners
and service providers.

@@

\subsection{Using the classification framework}

In order to demonstrate usage of our framework's constructs, we ranked our social machines
by their Alexa ranking, and performed a full repertory grid elicitation exercise over them.
This process involved ranking each social machine against the constructs, rating each construct
from 1 to 5. In total, 680 ranking were made, and the
results are illustrated in a grid and accompanying dendrograms for element and construct
similarity in Figure~\ref{dendrogram}.

>>>>

\section{Social machines: an introduction}
Once upon a time 'machines' were programmed by programmers and used by users. The success of the Web has changed this relationship: we now see configurations of people interacting with content and with each other, typified by social Web sites. Rather than drawing a line through such Web-based systems to separate the human and digital parts (as computer science has traditionally done), we can now draw a line around them  and  treat  each  such  compound  as  a  'social  machine'  -  a  machine in which the two aspects are seamlessly interwoven. This crucial transition in thinking acknowledges the reality of current socio-technical systems, and is essential to underpin any understanding of the science and engineering of their future development towards pervasive ecosystems of co-evolving social machines.

Essentially social machines can be characterized as assemblies of manually executed and machine-driven (as in 'automatized') services and the interaction of such services. A  traditional  database,  with  users  looking  up  records  independently  of  each  other  and  an  administrator responsible for the management of the content, has some of the right ingredients, but there is really no social element in this strict provider-consumer relationship. When we fill out a form (e.g., health information, birds spotted in my garden, new construction sites on my way to work) the systems supporting this activity are minimal forms of social machines  because the users  are  part  of  the  'social  computation', in this case the data creation and collection process facilitate through the site.  This social component  becomes  richer when  the database  is  curated by members  of  the broader community (e.g., Wikipedia) and when  the social network adds value implicitly (Amazon) or explicitly (Facebook) to the overall system through their individual or joint activities.

A number of prominent examples aside, today we still see a divide between conventional IT systems dedicated to data- and computation-intensive tasks, and Web $2.0$ sites offering some combination of well-known participatory features, in which user-generated content and the underlying social network evolve dynamically and hand-in-hand. However, as technology becomes more and more ubiquitous, many of the challenges we witness in any domain of our life, business, and society will soon require solutions that rely on both of these axes: a sophisticated  combination  of  data-intensive, complex automation and deep community involvement. This suggests the need for new types of systems to tackle these emerging challenges, and these systems will not be able to be built and used in a sustainable way without a through understanding of the science and engineering of (the continuum of) social machines.

As an early step towards achieving this principled understanding, we propose in this paper an initial outline for a classification framework for social machines. The primary aim of the framework is to identify and define the constructs to describe, study, and compare this expanding field of interdisciplinary research, as well as the increasing number of systems combining human and computational intelligence that are continuously being created in areas as diverse as traffic management, healthcare, media and journalism, or design and natural sciences.

The current version of our framework is the result of a knowledge elicitation process performed internally in our lab. Our goal is to present it to the broader community and extend and revise it based on their feedback in order to achieve a shared understanding of the basic notions and of the associated terminology, and to consensually lay out the structure of the design space that will enable a systematic study of existing and future social machines. We anticipate that the framework will be a useful tool for both researchers in social and computer sciences, and for developers and operators of social machines. Using a common conceptualization allows the former to become familiar with the landscape, and identify topics of research that so far have remained unexplored or under-explored. They are given a theoretical grounding to observe the effects of specific technical properties of a system configuration on social behavior, discover design and evolution patterns of such systems, learn about social network formation and dynamics, and devise incentive mechanisms to encourage wide participation. For developers and operators, the framework may inform the engineering of new systems in terms of critical features and community development.

This paper makes two main contributions:

\begin{itemize}
  \item We introduce the notion of 'social machines' and position it in connection to similar ideas and initiatives.
  \item We propose a set of constructs to capture the most important features of social machines and to compare existing instances thereof.
\end{itemize}

The remainder of this paper is organized as follows. In order to pin down what makes up a social machine, and establish the boundaries of this emerging field of research, we need to first understand the relationship between social machines and related topics such as human computation, collective intelligence, and crowdsourcing; Section \ref{sec:comparison} gives a summary of this comparative analysis. In Section \ref{sec:framework} we then present the framework in terms of the methodology used to devise it, the main clusters of constructs used for classification, and an example of using them to compare a selection of popular social machines via grid analysis. We conclude with an outlook of our future development activities, including details about the evaluation of the framework. In addition, we plan to engage with the broader community to set up a 'social machine' bringing together researchers from various disciplines who will contribute to the 'computation' of a joint and widely agreed classification framework for the field.

\section{Comparison with related fields}
\label{sec:comparison}
Since the invention of the Turing machine we have experienced a 'paradigm shift' in the usage of computers, moving away from purely calculative devices to facilitators of a wide range of human interactions. The emergence of 'Computer-supported Collaboration Work (CSCW)' \cite{grudin1994computer} is representative for the early days of this trend. CSCW addresses 'how collaborative activities and their coordination can be supported by means of computer systems' \cite{grudin1994computer}, whereas the initial concept evolved towards the more broader field of 'Computer-supported collaboration' (CSC).

The Web as a global platform for information access and sharing marked a second essential milestone; in particular, through principles and technologies promoted through Web $2.0$ and the Mobile Web. These developments led to amazing growth in terms of the amounts of content available online and mass participation. They are responsible for hundreds of millions of users all over the globe creating high-quality encyclopedias, publishing Terabytes of multimedia content, contributing to world-class software, and lively taking part in defining the agenda of many aspects of our society. This trend towards 'prosumerism' is finding more and more adopters in the public and private sector as well, as governments and enterprises not only become active in open initiatives, but encourage the participation of their customers and employees in taking decisions related to organizational management, product development, services offers, and policies. In this context, a number of terms are used to refer to the ways people interact with each other and with applications: 'wisdom of the crowds', 'collective intelligence', 'open innovation', 'crowdsourcing', 'human computation', and 'social computing'. They are related, but not synonymous with 'social machines', as explained in the following (see also Figure \ref{socialmachine}).

\begin{figure}[htb]
\begin{center}
\includegraphics[width=8cm]{img/socialmachinescope.png}
\caption{Social machines and related areas} \label{fig:social machine}
\end{center}
\end{figure}

Wisdom of the crowds \cite{surowiecki2005wisdom} refers to a principle for decision making that takes into account information and opinions of a group of people rather than individuals; the use of specific technologies, most notably Web $2.0$, has made it possible for such processes to be carried out at scales hardly conceivable in the past, and to involve highly diverse and geographically distributed participants. A similar concept, though broader scoped, is collective intelligence, defined in  \cite{malone2009harnessing} as 'groups of individuals doing things collectively that seem intelligent.' The main difference to what we refer to as 'social machines' is hidden in the the automation part of the latter. In a social machine, human and computational intelligence coalesce in order to achieve a given purpose. The overall setting may slightly vary in terms of the interaction patterns between the two types of services and the way their outcomes are used; however, where wisdom of the crowds and collective intelligence clearly place their focus on identifying the situations in which groups of people perform better than individuals, social machines are concerned with the study and realization of hybrid systems where the two types of components co-exist. As such, theoretical and empirical insights from the two related areas are useful to understand the dynamics of the social structures underlying such a system, but they are definitely not the only ingredient needed to build operational social machines. The question of how human and automatic services can be brought together to achieve optimal results, as well as the actual engineering exercise by which a system is developed, tested, and updated are equally important.

One of the direct consequences of the popularity of the wisdom of the crowds idea was a stronger investment worldwide in open innovation, which can be seen as an application of the concept to business environments, or, in the words of the authors, as a 'a paradigm that assumes that firms can and should use external ideas as well as internal ideas, and internal and external paths to market, as the firms look to advance their technology' \cite{chesbrough2003}. At a more general level, this kind of creativity could be leveraged in almost any domain which benefits from diversity, while in the same time ensuring in-time access of a potentially infinite pool of skills and resources not available before the advent of social computing technologies. The term crowdsourcing is typically associated with this larger collection of situations, in which 'a job traditionally performed by a designated agent [...] [is outsourced] to an undefined, generally large group of people in the form of an open call' \cite{howe2006crowdsourcing}. Human computation applies human processing power to tackle technical tasks that computers (still) find challenging \cite{von2009human}, typically in areas such as visual, audio, and natural language understanding. This sort of tasks are an important part of today's crowdsourcing projects landscape, in particular on so-called 'microtask' platforms such as Amazon's Mechanical Turk\footnote{\url{https://www.mturk.com/}} or CrowdFlower\footnote{\url{http://crowdflower.com/}}, which offer small financial rewards to an anonymous crowd engaged with atomic units of work in the range of seconds to minutes to complete.  The key difference between social machines and open innovation and crowdsourcing is again in the computational part. The novelty these two approaches bring in lies in using a much larger pool of human resources as traditional work environments \cite{quinn2011human}. There are clear socio-economic implications that their adopters need to deal with in order to optimally make use of this wealth of resources; technology may be needed to assist specific aspects of crowdsourcing projects, from evaluating and rewarding the results produced by the crowd to consolidating and aggregating them into a complete solution. A principled combination of human and computational capabilities is less in the focus of such systems and the technical means that support them, by contrast to social machines. In comparison to human computation, social machines cover a wider range of scenarios. Human computation is AI-centric and uses people to perform tasks that computers are not (yet) able to tackle (in terms of accuracy) \cite{quinn2011human}; by comparison, we see many successful examples of social machines in which the role of machines is rather to facilitate interactions within groups of people or communities of interest.

An analogous line of reasoning could be followed to point out the overlaps between social machines and social computing \cite{parameswaran2007research}. The latter is an area of computer science which refers systems that support 'the gathering, representation, processing, use, and dissemination of information that is distributed across social collectivities such as teams, communities, organizations, and markets' \cite{parameswaran2007research}. As such, compared to the general concept of 'Computer-supported collaboration', social computing puts a greater emphasis on the information management capabilities of groups and communities, and less on the way these capabilities emerge as a joint effort. This distinction is even stronger in the case of social machines, which look at the social and the technical components as equal and necessary partners, and study the ways they could be best combined to master the challenges of future socio-technical systems.

We now turn to a description of the main classification dimensions and features relevant to describe, study, and compare social machines.

\section{Classification of social machines}
\label{sec:framework}

\subsection{Methodology}
\label{sec:methodology}
The classification framework presented in this paper was created through knowledge elicitation ~\cite{knowledgeelicitation}. In particular we used the {\it repertory grid} elicitation
technique~\cite{kelly} in order to derive an initial set of {\it elements}, which represent instances of social machines, and {\it constructs}, which capture their most important characteristics.

In repertory grid elicitation a software tool is used to ask users to describe constructs
that differentiate between elements - for example,  prompting the user to create a construct that
differentiates between {\it GalaxyZoo} and {\it Facebook}, as two prominent example of social machines our classification framework should apply to. The user describes the opposing poles of the construct - in this case, the user may decide on 'For Science' and 'For Socialising' in order to capture the core distinction between the two system. The user then rates every element with value from $1$ to $5$ on this construct, where $1$ represents an element that is purely
'For Science', and $5$ one that mainly serves  'socializing' purposes. Elements can also be rated with values between the poles if they may correspond to both or neither. The repertory grid software then identifies elements that are the least differentiated when requesting new constructs, in an attempt to elicit as much knowledge as possible from the user. We applied this technique to create an initial set of examples of social machines and to classify them according to specific features. To do so, we asked then computer science researchers familiar with the field to create their own repertory grids, and populate the elements from their own knowledge, and created the constructs using the standard repertory grid elicitation technique described above. This exercise led to $10$ grids, the union of which comprised a total of $56$ unique elements (social machines)
and $117$ different constructs (classifying factors). As the aim of this initial phase was to understand how people perceive the notion of social machine and their most distinctive properties, we allowed the participants to choose the social machines they are familiar with and describe them in their own terms.

@@

While determining the intersection of elements was straightforward, the consolidation of the
constructs required a more thorough process. We manually grouped the constructs into rough clusters, based around the areas they cover. We examined each construct to determine which were equivalent, and whether we could re-word or subsume existing constructs to cover the same aspects. The aim of this exercise was twofold: to remove any redundancy and cut the classification space down to a manageable size, while ensuring that all constructs that were elicited were represented in the final set. This process involved four of the authors discussing the choices, and resulted in a consolidated set of $31$ constructs organized in four different clusters:

\begin{description}
  \item [Popularity]: this cluster mainly covered descriptive information about a given system, such as its maturity and number of users. For the purpose of our analysis we decided to discard the associated constructs, as they 
  \item [Tasks and purpose]
  \item [Motivation and incentives]
  \item [Technology]
\end{description}

The paper presents the current version of the framework. We envision a reiteration of the main clusters and constructs based on discussions with and feedback from various scientific audiences. We provide initial evidence of the usefulness of the framework as a tool to  examine commonalities and differences between social machines in Section \ref{sec:usage}, in which we applied the repertory grid elicitation technique to a set of $20$ social machines instances ranked by popularity according to the Alexa service.\footnote{\url{www.alexa.com/}} A more thorough evaluation will

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{|p{8cm}|}
\hline
{\bf Users and Roles} \\
\hline
Generality of Audience \\
Participant autonomy \\
Participant anonymity \\
Extent of hierarchical organisation of roles \\
Clear separation of roles/responsibilities among participants \\
\hline
{\bf Tasks} \\
\hline
Activities involve creative production of content \\
Activities involve subjective appraisal of content \\
Activities involve solving (a definable) computation task or set of problems\\
Tasks are domain-specific \\
The machine owner (service/platform) derives benefit from participation \\
Activities and tasks are pre-defined or participant-defined \\
Variation in types of contributions and tasks \\
\hline
{\bf Motivation} \\
\hline
Participants are intrinsically motivated:\\
\hspace{1cm} to gain/share knowledge \\
\hspace{1cm} to ``get something done'' \\
\hspace{1cm} to ``for fun/entertainment''\\
\hspace{1cm} to ``be social'' \\
\hspace{1cm} for the benefit of a specific group of people who need help \\
\hspace{1cm} for the benefit of society as a whole \\
Participation is motivated by extrinsic reward (payment, status etc) \\
\hline
{\bf Context(s) of Participation} \\
\hline
Participants' participate to fulfill needs of a role  \\
\hspace{1cm} work-related/professional role \\
\hspace{1cm} home/family related \\
\hspace{1cm} social role \\
\hspace{1cm} leisure/entertainment role \\
Participation is done via: \\
\hspace{1cm} mobile devices \\
\hspace{1cm} web browsers \\
\hspace{1cm} apps \\
\hspace{1cm} sensors/sensing (location sensing and wearable devices) \\
Participation is done in a mobile context \\
\hspace{1cm} Physical location is relevant to the service \\
\hline
\end{tabular}
\end{center}
\caption{Consolidated constructs of social machines.} \label{table:constructs}
\end{table}

@@

\subsubsection{Tasks, Problems and Activities}
The first set of constructs pertain to \emph{what} the participants
do, both individually within the system, and collectively as a whole.
Individual activities might include creation of content, subjective
appraisal of existing content, posing of problems, solving of
problems, andf so on.  The resulting computation, looked at from a
macroscopic whole, might be the identification of high-quality
artistic content, insight, or creative works, collective problem
solving, and so on.  Such tasks might be pre-defined by the system
designers, or brought to the system by participants.  Futher, these
tasks might be of a specific type or class, or of many varied types or
encompass a general class of activities.  To capture these
difeferences, we include a construct that differentiates machines with
pre-defined problem spaces or are open to participant-provided tasks
and activities.  Finally, we ask whether the benefit derived by
participants is distinct from that gained by machine/platform owners
and service providers.

@@

\subsection{Using the classification framework}
\label{sec:usage}
In order to demonstrate usage of our framework's constructs, we ranked our social machines
by their Alexa ranking, and performed a full repertory grid elicitation exercise over them.
This process involved ranking each social machine against the constructs, rating each construct
from 1 to 5. In total, $680$ ranking were made, and the
results are illustrated in a grid and accompanying dendrograms for element and construct
similarity in Figure~\ref{dendrogram}.
